Something about Performance, 2D Component of NEMO. (TBC)

### 2016-08-01 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d: update to intro

@@

The \psykal\ approach attempts to address the problems described in the
previous section. It separates code into three layers; the Algorithm
layer, the PSy layer and the Kernel layer. The approach has been
developed in the GungHo project~\cite{GungHo}, which is creating a new
Dynamical core for the Met Office, and its design has been influenced
by earlier work on OP2~\cite{OP2,PYOP2}.

@@

While the \psykal\ approach is general, we are currently applying it to
Atmosphere and Ocean models written in Fortran where domain
decomposition is typically performed in the latitude-longitude
direction, leaving columns of elements on each domain-decomposed
partition.

@@

%xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

Proliferation of lightweight cores is resulting in increased
complexity in the programming models required to achieve performance.
Combine with the complexity of e.g. a model simulating circulation in
the global ocean. Significant challenge in software engineering.
Propose an approach for meeting this challenge by separating the
computational science demands from the {\it e.g.} oceanographic
demands.

\begin{figure}
\centering
\caption{The \psykal\ Approach.}
\end{figure}

While encapsulation makes it easier to produce more robust software,
it can often come with a performance penalty. In this work we
investigate in detail the hardware and compiler dependencies of this
penalty and the steps required to remove it.

\subsection{The `nemolite2d' Program}

For this work we have used a program "nemolite2d" developed by
ourselves.  “nemolite2d” is a vertically-averaged version of NEMO
(Nucleus for European Modelling of the Ocean~\cite{nemo}), which only keeps its
dynamical part. The whole model system is represented through one
continuity equation \eqref{continuity} (for the update of the sea-surface height) and two
vertically-integrated momentum equations \eqref{momentum} (for the two velocity
components, respectively).

@@

where $\zeta$ and $\vec{U}$ represent the sea surface height and horizontal velocity vector
respectively. $h$ is the total water depth, $\Omega$ is the Earth rotation velocity vector.
$g$ is the gravity. $\nu$ is the kinematic viscosity coefficient.

The external forcing includes surface wind stress, bottom friction,
and open boundary barotropic forcing. A lateral slip boundary
condition is applied along the coast lines. The open boundary
conditions can be set as a clipped or a Flather's radiation
condition~\cite{flather76}. The bottom friction takes a semi-implicit
form for the sake of model stability. As done in the original version
of NEMO, a constant or Smagorinsky horizontal viscosity coefficient is
used for the horizontal viscosity term.

The traditional Arakawa C structured grid is employed here for the
discretisation of the computational domain. A two-dimensional integer
array is used to identify the different parts of the computational
domain; it has the value of one for ocean, zero for land and minus
one for ocean cells outside of the computational domain. This array
serves as a mark to identify ocean cells, land cells, solid boundaries
and open boundaries.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal\ version of nemolite2d retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the \psykal\ version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal\ version of nemolite2d, we made the
kernels as fine-grained as possible. This resulted in eight distinct
kernels, each of which updates a single field at a single point (since
we have chosen to use point-wise kernels). With a little bit of
tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
gives us a single PSy-layer routine which consists of applying each of
the kernels to all of the points requiring an update on the model
mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
routine then contains a doubly-nested loop around each kernel call, as
indicated by the pseudo-code in
Figure~\ref{FIG_psykal_nemolite2d_structure}.

\begin{figure}
\centering
%\includegraphics[width=85mm]{psykal_nemolite2d}
\caption{A schematic of the vanilla \psykal\ version of the nemolite2d code.}
\label{FIG_psykal_nemolite2d_structure}
\end{figure}

As with any full oceanographic model, boundary conditions must be
applied at the edges of the model domain. Since nemolite2d applies
external boundary conditions ({\it e.g.} barotropic forcing), this is
done via user-supplied kernels.

@@

We first describe the code transformations performed for the serial
version of nemolite2d.  We then move on to the construction of
parallel versions of the code using OpenMP and OpenACC. Again, we
describe the key steps we have taken in this process in order to
maximise the performance of the code.

\subsection{Transformations of Serial Nemolite2D}

@@

\begin{table}[!t]
\caption{The performance profile of the original and \psykal\ versions
  of nemolite2d on the Intel Ivy Bridge CPU (for 2000 time-steps of the
  $128^{2}$ domain and the Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
\multicolumn{2}{c}{Final \psykal} \\
        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
\hline
Momentum    & 1.98  & 72.6  & 4.05  & 79.5 & 2.09 & 75.3 \\
Time-update & 0.40  & 14.6  & 0.41  & 8.1  & 0.29 & 10.6 \\
BCs         & 0.25  & 9.1   & 0.29  & 5.7  & 0.27 & 9.9  \\
Continuity  & 0.10  & 3.7   & 0.33  & 6.6  & 0.11 & 4.1  \\
\hline
\end{tabular}
\end{table}

@@

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the nemolite2d kernels contain conditional statements. These
statements are there to check whether {\it e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
ftn-6375 ftn: VECTOR File = nemolite2d.f90,
 Line = 448 
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all field-data arrays with extents greater than
strictly required.  We were then able to safely add the {\tt safe\_address}
before all of the loops where the Cray compiler indicated it might be
useful (the Momentum loops and some of the BC loops).

@@

It turns out that the Cray-compiled binaries of both the original and
\psykal\ versions of nemolite2d perform considerably less well than
their Intel-compiled counterparts. Comparison of the diagnostic output
from each of the compilers revealed that while the Intel compiler was
happy to vectorise the Momentum loops, the Cray compiler was choosing
not to:
\begin{verbatim}
   99.          do ji = 2, M-1, 1
 A loop starting at line 99 was blocked 
 with block size 256.

@@

\subsection{Construction of OpenMP-Parallel Nemolite2D}
\label{sec_omp_steps}

For this part of the work we began with the optimal \psykal\ version of
the code, as obtained after applying the various transformations
described in the previous section. As with the transformations of the
serial code, our purpose here is to determine the functionality
required of the PSyclone code-generation system.

@@

The simplest possible OpenMP-parallel implementation consists of
parallelising each loop nest in the PSy layer. This was done by
inserting an OpenMP PARALLEL DO directive before each loop nest so
that the iterations of the outermost or $j$ loop (over the y-dimension
of the model domain) are shared out amongst the OpenMP threads. This
leaves the innermost ($i$) loop available for SIMD vectorisation by
the compiler.

The loop nest dealing with the application of the Flather boundary
condition to the y-component of velocity has a loop-carried dependency
in $j$ which appears to prevent its being executed in parallel. This was
therefore left unchanged and executed on thread 0 only.

@@

By default, the OpenMP END DO directive includes an implicit barrier,
thus causing all threads to wait until the slowest has completed the
preceeding loop. Such synchronisation limits performance at larger
thread counts and, for the nemolite2d code, is frequently unecessary.
{\it E.g.} if a kernel does not make use of the results of a
preceeding kernel call then there is clearly no need for threads to
wait between the two kernels.

@@

We first consider the performance of the code in serial and examine
the effects of the transformations described in
Section~\ref{sec_methodology}. Once we have arrived at an optimal form
for the serial version of nemolite2d we then investigate its parallel
performance on both CPU- and GPU-based systems.

@@

In Figure~\ref{FIG_orig_perf_summary} we plot the serial performance
of the original version of the nemolite2d code for the range of
compilers and CPUs considered here. Unlike the Shallow
code~\cite{shallow_psykal}, the original version of nemolite2d has not
been optimised. Although it is still a single source file it is, in
common with NEMO itself, logically structured with separate
subroutines performing different parts of the physics within each time
step. This structuring and the heavy use of conditional statements
favour the Intel compiler which significantly out-performs both the
Gnu and Cray compilers. Only when the problem size falls out of cache
does the performance gap begin to narrow. The reason for the
performance deficit of the Cray-compiled binary comes down to (a lack
of) SIMD vectorisation, an issue that we explore below.

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{orig_summary}
\caption{Summary of the performance of the original version of the 
nemolite2d code on 
the range of compilers and CPU-based systems under consideration.}
\label{FIG_orig_perf_summary}
\end{figure}

Moving now the to the \psykal\ version of nemolite2d,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest \psykal\ version for each of the compiler/CPU combinations.
While the Intel compiler still produces the best-performing binary,
the Cray-compiled binary is now a very close second. In fact, the
performance of both the Gnu- and Cray-compiled \psykal\ versions is
generally significantly greater than that of their respective original
versions. We also note that best absolute performance (in terms of
grid points processed per second) with any compiler is obtained with
the $256^2$ domain.

\begin{figure}[!t]
\centering
\includegraphics[width=7.5cm]{best_psykal_summary}
\caption{Summary of the best performance achieved by any \psykal\ 
version of nemolite2d with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

Figure~\ref{FIG_slowdown_summary} plots the percentage difference
between the performance of the original and the best \psykal\ versions
of nemolite2d for each compiler/CPU combination. This shows that it is
only the Intel-compiled binary running the $64^2$ domain that is
slower with the \psykal\ version of the code (and then only by some
3\%). For all other points in the space, the optimised \psykal\ version
of the code performs better. The results for the Cray compiler are
however somewhat sqewed by the fact that it did not SIMD vectorise key
parts of the original version (see below).

@@

Having show that we can recover, and often improve upon, the
performance of the original version of nemolite2d, the next logical
step is to examine the necessary code transformations in detail.  We
do this for the $256^{2}$ case since this fits within cache on the Ivy
Bridge CPUs we are using here.  Table~\ref{TABLE_opt_breakdown} shows
detailed performance figures for this case after each transformation has been
applied to the code. The same data is visualised in
Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal\
version for a given compiler/CPU is given relative to the performance
of the original with the same compiler/CPU.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=93mm]{opt_stages_256}
\caption{Performance of the \psykal\ version of nemolite2d for the
  $256^{2}$ domain at each stage of optimisation. The first (black)
  bar of each cluster gives the performance of the original version of
  nemolite2d for that compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

@@

We now turn to transformations related to parallelisation of the
nemolite2d code; the introduction of OpenMP and OpenACC directives. In
keeping with the \psykal\ approach we do not modify either the
Algorithm- or Kernel-layer code.  Any code changes are restricted to
either the PSy (middle) layer or the underlying library that manages
{\it e.g.} the construction of field objects.

@@

In order to quantify the scaling behaviour of the different versions
of nemolite2d with the different compilers/run-time environments, we
also plot the parallel efficiency in
Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
and~\ref{FIG_cray_omp_256} (dashed lines and open symbols). We define
parallel efficiency (\%), $E(n)$, on a number of threads, $n$, as:
\[
E(n) = 100  \frac{P(n)}{ n P(1)}
\]
where $P(m)$ is the performance of the code on $m$ threads. For a
perfect, linearly-scaling code, $E(n)$ will be 100\%.

Since the space to explore consists of three different compilers, six
different domain sizes, five stages of optimisation and six different
thread counts we can only consider a slice through it in what follows.
In order to inform our choice of domain size,
Figure~\ref{FIG_omp_domain_size} shows the scaling behaviour of the
most performant, Intel-compiled version of nemolite2d. Although it is
common for production runs of NEMO to use MPI sub-domains of $20
\times 20$ ({\bf do we have/need a reference for that?}), it is clear
from Figure~\ref{FIG_omp_domain_size} that the limited quantity of
parallelism in the $32 \times 32$ domain inhibits scaling. Therefore,
to fully test the performance of our OpenMP implementations we have
chosen to examine results for the $256 \times 256$ domain and these
are shown in Figures~\ref{FIG_gnu_omp_256}, \ref{FIG_intel_omp_256}
and \ref{FIG_cray_omp_256}.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_problem_size}
\caption{The scaling behaviour of the most performant OpenMP-parallel
  version of \psykal\ nemolite2d for the full range of domain sizes
  considered on the CPU. Results are for the Intel compiler on a
  single Intel Ivy Bridge socket. The corresponding parallel
  efficiencies are shown using open symbols and dashed lines. The
  24-thread runs employed hyperthreading.}
\label{FIG_omp_domain_size}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_gnu}
\caption{Performance of the OpenMP-parallel version of \psykal\
  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
  socket. The corresponding parallel efficiencies are shown using open
  symbols and dashed lines. The 24-thread runs employed hyperthreading
  and the optimal OpenMP schedule is given in parentheses.}
\label{FIG_gnu_omp_256}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_intel}
\caption{Performance of the OpenMP-parallel version of \psykal\
  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
  socket.  The corresponding parallel efficiencies are shown using
  open symbols and dashed lines. The 24-thread runs employed
  hyperthreading and the optimal OpenMP schedule is given in
  parentheses.}
\label{FIG_intel_omp_256}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_cray}
\caption{Performance of the OpenMP-parallel version of \psykal\
  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
  socket.  The corresponding parallel efficiencies are shown using
  open symbols and dashed lines. The 24-thread runs employed
  hyperthreading and the optimal OpenMP schedule is given in
  parentheses.}
\label{FIG_cray_omp_256}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{1and2sockets}
\caption{Performance of the OpenMP-parallel version of \psykal\
  nemolite2d on one and two sockets of Intel Ivy Bridge. The 24-thread
  runs on a single socket used hyperthreading and the two-socket runs
  had the threads shared equally between the sockets.}
\label{FIG_omp_2socks}
\end{figure}

@@

The introduction of OpenACC directives proved to be simpler than
producing the OpenMP version of nemolite2d {\bf - Jeremy to clarify
  whether this is true}.  Consequently we only show the performance of
{\bf xxx} versions of the code in Figure~\ref{FIG_gpu_opt_stages}.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{gpu_opt_stages}
\caption{Performance of the various GPU implementations of nemolite2d.}
\label{FIG_gpu_opt_stages}
\end{figure}

In Figure~\ref{FIG_cpu_cf_gpu} we compare the absolute performance of
the OpenMP and OpenACC implementations of nemolite2d across the range
of problem sizes considered. The OpenMP figures are the maximum
performance obtained from a whole Intel Ivy Bridge socket by any
version of the code on any number of threads for a given
compiler/run-time. For the smallest domain size ($64^2$) the OpenMP
version significantly outperforms the GPU because there is
insufficient parallelism to fully utilise the GPU and one time-step
takes only 80~$\mu$s. The execution of a single time-step is then
dominated by the time taken to launch a kernel on the GPU rather than
the execution of the kernel itself.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{cpu_cf_gpu}
\caption{Performance of the best OpenMP-parallel version of \psykal\
  nemolite2d (on a single Intel Ivy Bridge socket) compared with the best GPU implementation. {\bf This should really be best OpenACC version as that's what PSyclone can support.}}
\label{FIG_cpu_cf_gpu}
\end{figure}

@@

Although we have investigated how the performance of the \psykal
version of nemolite2d compares with that or the original, we have not
addressed how efficient the original actually is. In order to do so we
use the Roofline Model~\cite{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound. To do so we
follow the approach suggested in~\cite{para_pearls} and use the
STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks in order to
get the upper bounds on the memory bandwidth and floating-point
operations per second (FLOPS) for the test hardware. Since we are
using an Intel Ivybridge CPU we used the Intel Math Kernel Library
implementation of LINPACK.

>>>>

In a previous paper~\cite{shallow_psykal} we introduced a possible
approach to tackling this problem which we term \psykal\ (discussed
below). In that work we considered the implications for serial
performance of the extensive code re-structuring required by the
approach when applied to a shallow-water model. We found that although
the re-structuring did initially incur a sizeable performance penalty,
it was possible to transform the resulting code to recover performance
while obeying the \psykal\ separation of concerns. In this work we
move to looking at parallel performance within the \psykal\ approach.

@@

The \psykal\ approach attempts to address the problems described in
the previous section. It separates code into three layers; the
Algorithm layer, the PSy layer and the Kernel layer. The approach has
been developed in the GungHo project~\cite{GungHo}, which is creating
a new Dynamical core for the UK Met Office, and its design has been
influenced by earlier work on OP2~\cite{OP2, PYOP2}.

@@

While the \psykal\ approach is general, we are currently applying it to
Atmosphere and Ocean models written in Fortran where domain
decomposition is typically performed in the latitude-longitude
dimension, leaving columns of elements on each domain-decomposed
partition.

@@

\subsection{The `NEMOLite2D' Program}

For this work we have used a program, `NEMOLite2D,' developed by
ourselves.  NEMOLite2D is a vertically-averaged version of NEMO
(Nucleus for European Modelling of the Ocean~\cite{nemo}), retaining
only its dynamical part. The whole model system is represented through
one continuity equation \eqref{continuity} (for the update of the
sea-surface height) and two vertically-integrated momentum equations
\eqref{momentum} (for the two velocity components, respectively).

@@

where $\zeta$ and $\vec{U}$ represent the sea-surface height and
horizontal velocity vector, respectively. $h$ is the total water
depth, $\Omega$ is the Earth rotation velocity vector.  $g$ is the
acceleration due to gravity. $\nu$ is the kinematic viscosity coefficient.

The external forcing includes surface wind stress, bottom friction,
and open-boundary barotropic forcing. A lateral-slip boundary
condition is applied along the coast lines. The open boundary
condition can be set as a clipped or Flather's radiation
condition~\cite{flather76}. The bottom friction takes a semi-implicit
form for the sake of model stability. As done in the original version
of NEMO, a constant or Smagorinsky horizontal viscosity coefficient is
used for the horizontal viscosity term.

The traditional Arakawa C structured grid is employed here for the
discretisation of the computational domain. A two-dimensional integer
array is used to identify the different parts of the computational
domain; it has the value of one for ocean, zero for land and minus
one for ocean cells outside of the computational domain. This array
enables the identification of ocean cells, land cells, solid boundaries
and open boundaries.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal\ version of NEMOLite2D retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the \psykal\ version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal\ version of NEMOLite2D, we made the
kernels as fine-grained as possible. This resulted in eight distinct
kernels, each of which updates a single field at a single point (since
we have chosen to use point-wise kernels). With a little bit of
tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
gives us a single PSy-layer routine which consists of applying each of
the kernels to all of the points requiring an update on the model
mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
routine then contains a doubly-nested loop around each kernel call.

\begin{figure}
\centering
\begin{verbatim}
DO istep = 1, nsteps
  call invoke( continuity(ssha_t, ...), &
               momentum_u(ua, un, ...), &
               ...,                     &
               next_sshu(sshn_u, ...),  &
               next_sshv(sshn_v, ...) )
  call model_write(istep, sshn_t, un, vn)
END DO
\end{verbatim}
\caption{A schematic of the top-level of the \psykal\ version of the
 NEMOLite2D code.}
\label{FIG_psykal_nemolite2d_structure}
\end{figure}

As with any full oceanographic model, boundary conditions must be
applied at the edges of the model domain. Since NEMOLite2D applies
external boundary conditions ({\it e.g.} barotropic forcing), this is
done via user-supplied kernels.

@@

We first describe the code transformations performed for the serial
version of NEMOLite2D.  We then move on to the construction of
parallel versions of the code using OpenMP and OpenACC. Again, we
describe the key steps we have taken in this process in order to
maximise the performance of the code. In both cases our aim is to
identify those transformations which must be supported by any tool
which seeks to auto-generate a performant PSy layer.

\subsection{Transformations of Serial NEMOLite2D}

@@

\begin{table}[!t]
\caption{The performance profile of the original and \psykal\ versions
  of NEMOLite2D on the Intel Ivy Bridge CPU (for 2000 time-steps of the
  $128^{2}$ domain and the Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
\multicolumn{2}{c}{Final \psykal} \\
        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
\hline
Momentum    & 1.98  & 72.6  & 4.05  & 79.5 & 2.09 & 75.3 \\
Time-update & 0.40  & 14.6  & 0.41  & 8.1  & 0.29 & 10.6 \\
BCs         & 0.25  & 9.1   & 0.29  & 5.7  & 0.27 & 9.9  \\
Continuity  & 0.10  & 3.7   & 0.33  & 6.6  & 0.11 & 4.1  \\
\hline
\end{tabular}
\end{table}

@@

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the NEMOLite2D kernels contain conditional statements. These
statements are there to check whether {\it e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
ftn-6375 ftn: VECTOR File = nemolite2d.f90,
 Line = 448 
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all field-data arrays with extents greater than
strictly required.  We were then able to safely add the {\tt safe\_address}
before all of the loops where the Cray compiler indicated it might be
useful (the Momentum loops and some of the BC loops).

@@

It turns out that the Cray-compiled binaries of both the original and
\psykal\ versions of NEMOLite2D perform considerably less well than
their Intel-compiled counterparts. Comparison of the diagnostic output
from each of the compilers revealed that while the Intel compiler was
happy to vectorise the Momentum loops, the Cray compiler was choosing
not to:
\begin{verbatim}
   99.          do ji = 2, M-1, 1
 A loop starting at line 99 was blocked 
 with block size 256.

@@

\subsection{Construction of OpenMP-Parallel NEMOLite2D}
\label{sec_omp_steps}

For this part of the work we began with the optimal \psykal\ version of
the code, as obtained after applying the various transformations
described in the previous section. As with the transformations of the
serial code, our purpose here is to determine the functionality
required of a tool that seeks to generate the PSy layer.

@@

The simplest possible OpenMP-parallel implementation consists of
parallelising each loop nest in the PSy layer. This was done by
inserting an OpenMP PARALLEL DO directive before each loop nest so
that the iterations of the outermost or $j$ loop (over the latitude
dimension of the model domain) are shared out amongst the OpenMP
threads. This leaves the innermost ($i$) loop available for SIMD
vectorisation by the compiler.

The loop nest dealing with the application of the Flather boundary
condition to the y-component of velocity has a loop-carried dependency
in $j$ which appears to prevent its being executed in
parallel.\footnote{Only once this work was complete did we establish
  that boundary conditions are enforced such that it can be executed
  in parallel.}  This was therefore left unchanged and executed on
thread 0 only.

@@

By default, the OpenMP END DO directive includes an implicit barrier,
thus causing all threads to wait until the slowest has completed the
preceeding loop. Such synchronisation limits performance at larger
thread counts and, for the NEMOLite2D code, is frequently unecessary.
{\it E.g.} if a kernel does not make use of the results of a
preceeding kernel call then there is clearly no need for threads to
wait between the two kernels.

@@

We first consider the performance of the code in serial and examine
the effects of the transformations described in
Section~\ref{sec_methodology}. Once we have arrived at an optimal form
for the serial version of NEMOLite2D we then investigate its parallel
performance on both CPU- and GPU-based systems.

@@

In Figure~\ref{FIG_orig_perf_summary} we plot the serial performance
of the original version of the NEMOLite2D code for the range of
compilers and CPUs considered here. Unlike the Shallow
code~\cite{shallow_psykal}, the original version of NEMOLite2D has not
been optimised. Although it is still a single source file it is, in
common with NEMO itself, logically structured with separate
subroutines performing different parts of the physics within each time
step. This structuring and the heavy use of conditional statements
favour the Intel compiler which significantly out-performs both the
Gnu and Cray compilers. Only when the problem size falls out of cache
does the performance gap begin to narrow. The reason for the
performance deficit of the Cray-compiled binary comes down to (a lack
of) SIMD vectorisation, an issue that we explore below.

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{orig_summary}
\caption{Summary of the performance of the original version of the 
NEMOLite2D code on 
the range of compilers and CPU-based systems under consideration.}
\label{FIG_orig_perf_summary}
\end{figure}

Moving now the to the \psykal\ version of NEMOLite2D,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest \psykal\ version for each of the compiler/CPU combinations.
While the Intel compiler still produces the best-performing binary,
the Cray-compiled binary is now a very close second. In fact, the
performance of both the Gnu- and Cray-compiled \psykal\ versions is
generally significantly greater than that of their respective original
versions. We also note that best absolute performance (in terms of
grid points processed per second) with any compiler is obtained with
the $256^2$ domain.

\begin{figure}[!t]
\centering
\includegraphics[width=7.5cm]{best_psykal_summary}
\caption{Summary of the best performance achieved by any \psykal\ 
version of NEMOLite2D with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

Figure~\ref{FIG_slowdown_summary} plots the percentage difference
between the performance of the original and the best \psykal\ versions
of NEMOLite2D for each compiler/CPU combination. This shows that it is
only the Intel-compiled binary running the $64^2$ domain that is
slower with the \psykal\ version of the code (and then only by some
3\%). For all other points in the space, the optimised \psykal\ version
of the code performs better. The results for the Cray compiler are
however somewhat sqewed by the fact that it did not SIMD vectorise key
parts of the original version (see below).

@@

Having show that we can recover, and often improve upon, the
performance of the original version of NEMOLite2D, the next logical
step is to examine the necessary code transformations in detail.  We
do this for the $256^{2}$ case since this fits within cache on the Ivy
Bridge CPUs we are using here.  Table~\ref{TABLE_opt_breakdown} shows
detailed performance figures for this case after each transformation has been
applied to the code. The same data is visualised in
Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal\
version for a given compiler/CPU is given relative to the performance
of the original with the same compiler/CPU.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=93mm]{opt_stages_256}
\caption{Performance of the \psykal\ version of NEMOLite2D for the
  $256^{2}$ domain at each stage of optimisation. The first (black)
  bar of each cluster gives the performance of the original version of
  NEMOLite2D for that compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

@@

We now turn to transformations related to parallelisation of the
NEMOLite2D code; the introduction of OpenMP and OpenACC directives. In
keeping with the \psykal\ approach we do not modify either the
Algorithm- or Kernel-layer code.  Any code changes are restricted to
either the PSy (middle) layer or the underlying library that manages
{\it e.g.} the construction of field objects.

@@

In order to quantify the scaling behaviour of the different versions
of NEMOLite2D with the different compilers/run-time environments, we
also plot the parallel efficiency in
Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
and~\ref{FIG_cray_omp_256} (dashed lines and open symbols). We define
parallel efficiency (\%), $E(n)$, on a number of threads, $n$, as:
\[
E(n) = 100  \frac{P(n)}{ n P(1)}
\]
where $P(m)$ is the performance of the code on $m$ threads. For a
perfect, linearly-scaling code, $E(n)$ will be 100\%.

Since the space to explore consists of three different compilers, six
different domain sizes, five stages of optimisation and six different
thread counts we can only consider a slice through it in what follows.
In order to inform our choice of domain size,
Figure~\ref{FIG_omp_domain_size} shows the scaling behaviour of the
most performant, Intel-compiled version of NEMOLite2D. Although it is
common for production runs of NEMO to use MPI sub-domains of $20
\times 20$ ({\bf do we have/need a reference for that?}), it is clear
from Figure~\ref{FIG_omp_domain_size} that the limited quantity of
parallelism in the $32 \times 32$ domain inhibits scaling. Therefore,
to fully test the performance of our OpenMP implementations we have
chosen to examine results for the $256 \times 256$ domain and these
are shown in Figures~\ref{FIG_gnu_omp_256}, \ref{FIG_intel_omp_256}
and \ref{FIG_cray_omp_256}.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_problem_size}
\caption{The scaling behaviour of the most performant OpenMP-parallel
  version of \psykal\ NEMOLite2D for the full range of domain sizes
  considered on the CPU. Results are for the Intel compiler on a
  single Intel Ivy Bridge socket. The corresponding parallel
  efficiencies are shown using open symbols and dashed lines. The
  24-thread runs employed hyperthreading.}
\label{FIG_omp_domain_size}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_gnu}
\caption{Performance of the OpenMP-parallel version of \psykal\
  NEMOLite2D for the Gnu compiler on a single Intel Ivy Bridge
  socket. The corresponding parallel efficiencies are shown using open
  symbols and dashed lines. The 24-thread runs employed hyperthreading
  and the optimal OpenMP schedule is given in parentheses.}
\label{FIG_gnu_omp_256}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_intel}
\caption{Performance of the OpenMP-parallel version of \psykal\
  NEMOLite2D for the Intel compiler on a single Intel Ivy Bridge
  socket.  The corresponding parallel efficiencies are shown using
  open symbols and dashed lines. The 24-thread runs employed
  hyperthreading and the optimal OpenMP schedule is given in
  parentheses.}
\label{FIG_intel_omp_256}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_cray}
\caption{Performance of the OpenMP-parallel version of \psykal\
  NEMOLite2D for the Cray compiler on a single Intel Ivy Bridge
  socket.  The corresponding parallel efficiencies are shown using
  open symbols and dashed lines. The 24-thread runs employed
  hyperthreading and the optimal OpenMP schedule is given in
  parentheses.}
\label{FIG_cray_omp_256}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{1and2sockets}
\caption{Performance of the OpenMP-parallel version of \psykal\
  NEMOLite2D on one and two sockets of Intel Ivy Bridge. The 24-thread
  runs on a single socket used hyperthreading and the two-socket runs
  had the threads shared equally between the sockets.}
\label{FIG_omp_2socks}
\end{figure}

@@

The introduction of OpenACC directives proved to be simpler than
producing the OpenMP version of NEMOLite2D {\bf - Jeremy to clarify
  whether this is true}.  Consequently we only show the performance of
{\bf xxx} versions of the code in Figure~\ref{FIG_gpu_opt_stages}.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{gpu_opt_stages}
\caption{Performance of the various GPU implementations of NEMOLite2D.}
\label{FIG_gpu_opt_stages}
\end{figure}

In Figure~\ref{FIG_cpu_cf_gpu} we compare the absolute performance of
the OpenMP and OpenACC implementations of NEMOLite2D across the range
of problem sizes considered. The OpenMP figures are the maximum
performance obtained from a whole Intel Ivy Bridge socket by any
version of the code on any number of threads for a given
compiler/run-time. For the smallest domain size ($64^2$) the OpenMP
version significantly outperforms the GPU because there is
insufficient parallelism to fully utilise the GPU and one time-step
takes only 80~$\mu$s. The execution of a single time-step is then
dominated by the time taken to launch a kernel on the GPU rather than
the execution of the kernel itself.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{cpu_cf_gpu}
\caption{Performance of the best OpenMP-parallel version of \psykal\
  NEMOLite2D (on a single Intel Ivy Bridge socket) compared with the best GPU implementation. {\bf This should really be best OpenACC version as that's what PSyclone can support.}}
\label{FIG_cpu_cf_gpu}
\end{figure}

@@

Although we have investigated how the performance of the \psykal
version of NEMOLite2D compares with that or the original, we have not
addressed how efficient the original actually is. In order to do so we
use the Roofline Model~\cite{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound. To do so we
follow the approach suggested in~\cite{para_pearls} and use the
STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks in order to
get the upper bounds on the memory bandwidth and floating-point
operations per second (FLOPS) for the test hardware. Since we are
using an Intel Ivybridge CPU we used the Intel Math Kernel Library
implementation of LINPACK.
